PLAE enables quick and easy searching across one million ocular transcriptomes

### 2021-10-25 manuscript_00.Rmd
text updates

date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: massive_integrated_eye_scRNA.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "PURPOSE: To create a high performance reactive web application to query gene expression across cell type, species, study, and other factors. METHODS: We wrote the Shiny web application PLAE (https://plae.nei.nih.gov) and tailored the structure of the underlying data (sc Eye in a Disk, scEiaD) to display scEiaD. RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes., XX cell types, XX ocular tissues, XX body tissues. As a test of the value of this unified pan-eye dataset, we show XX. CONCLUSION: The PLAE v0.80 web app serves the pan-ocular and body dataset, scEiaD. This offers the eye community a powerful and quick means to test hypotheses on gene and transcript expression across 54 body and 19 eye tissues. "
keywords: "RNA-seq, lens, cornea, retina, RPE, Snakemake, web, app"
---

@@

# (sup?) Table 1
Studies in scEiaD

# Figure 1 Distribution of Tissues and Cell Types in scEiaD
```{r, fig.width=8, fig.height=5}
# organ
meta_filter <- meta_filter %>% mutate(organism = factor(organism, levels = meta_filter$organism %>% unique()))
a <- meta_filter %>%
  mutate(Organ = case_when(Organ == 'Eye' ~ 'Eye',
                           TRUE ~ 'Body'),
         Tissue = gsub('\\.|_', ' ', Tissue),
         Tissue = gsub("RPE$","RPE-Choroid", Tissue),
         Tissue = gsub(" and Aorta","", Tissue)) %>% 
  filter(Organ == 'Body') %>% 
  group_by(Organ, Tissue,organism, study_accession) %>%
  summarise(Count = n()) %>%
  ungroup() %>% 
  mutate(organism = factor(organism, levels = meta_filter$organism %>% unique())) %>% 
  ggplot(aes(x=Tissue,y=Count, fill = organism, group = organism)) +
  ylab('Number of Cells') +
  geom_bar(stat='identity', position = 'dodge') +
  cowplot::theme_cowplot() +
  coord_flip() +
  scale_y_continuous(breaks = c(2500, 5000, 7500, 10000, 12500, 84000)) +
  ggbreak::scale_y_break(c(13000, 82000)) +
  scale_fill_discrete(drop=TRUE,
                      limits = levels(meta_filter$organism)) +
  theme(legend.position = "none")

# eye studies
b <-  meta_filter %>%
  mutate(Organ = case_when(Organ == 'Eye' ~ 'Eye',
                           TRUE ~ 'Body'),
         Tissue = gsub('\\.|_', ' ', Tissue),
         Tissue = gsub("RPE$","RPE-Choroid", Tissue),
         Tissue = case_when(Tissue == "Choroid" ~ 'RPE-Choroid',
                            TRUE ~ Tissue),
         Tissue = gsub(" and Aorta","", Tissue)) %>% 
  filter(Organ == 'Eye') %>% 
  group_by(Organ, Tissue,organism) %>%
  summarise(Count = n()) %>%
  ungroup() %>% 
  ggplot(aes(x=Tissue,y=Count, fill = organism, group = organism)) +
  ylab('Number of Cells') +
  geom_bar(stat='identity', position = 'dodge') +
  cowplot::theme_cowplot() +
  coord_flip() +
  theme(legend.position = "none")
# eye studies samples
c <-  meta_filter %>%
  mutate(Organ = case_when(Organ == 'Eye' ~ 'Eye',
                           TRUE ~ 'Body'),
         Tissue = gsub('\\.|_', ' ', Tissue),
         Tissue = gsub("RPE$","RPE-Choroid", Tissue),
         Tissue = gsub(" and Aorta","", Tissue)) %>% 
  filter(Organ == 'Eye') %>% 
  select(Organ, Tissue,organism, study_accession) %>%
  unique() %>%
  group_by(Organ, Tissue, organism) %>% 
  summarise(Count = n()) %>%
  ungroup() %>% 
  ggplot(aes(x=Tissue,y=Count, fill = organism, group = organism)) +
  ylab('Number of Studies') +
  geom_bar(stat='identity', position = 'dodge') +
  cowplot::theme_cowplot() +
  coord_flip() +
  theme(legend.position = "none")

# retina cell types
d <- meta_filter %>% filter(Organ == 'Eye') %>% 
  group_by(CellType_predict, study_accession, organism) %>%
  summarise(Count = n()) %>% filter(Count > 10) %>% ungroup() %>% 
  ggplot(aes(x=CellType_predict,y=Count, fill = organism, group = organism)) +
  ylab('Number of Cells') +
  geom_bar(stat='identity', position = 'dodge') +
  cowplot::theme_cowplot() +
  coord_flip()

plot_grid(a,b, c, d)
```

@@

## Data difficult to access

## Cross comparisons difficult
But important as seeing trends trends across different groups strongly demonstrates that 
the signal is biological, not technical.

@@

## PLAE contains 10 fold more cells and studies than the closest comparable web app

>>>>

date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: zotero_plaeApp.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "PURPOSE: To create a high performance reactive web application to query gene expression across cell type, species, study, and other factors. METHODS: We wrote the Shiny web application PLAE (https://plae.nei.nih.gov) and tailored the structure of the underlying data (sc Eye in a Disk, scEiaD) to display scEiaD. RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes., XX cell types, XX ocular tissues, XX body tissues. As a test of the value of this unified pan-eye dataset, we show XX. CONCLUSION: The PLAE v0.80 web app serves the pan-ocular and body dataset, scEiaD. This offers the eye community a powerful and quick means to test hypotheses on gene and transcript expression across 54 body and 19 eye tissues. "
keywords: "RNA-seq, lens, cornea, retina, RPE, Snakemake, web, app"
---

@@

# (sup?) Table 1
Studies in scEiaD
```{r}
meta_filter %>% group_by(Citation, PMID, study_accession) %>% summarise(Count = n()) %>% flextable()
```

# Figure 1 Distribution of Tissues and Cell Types in scEiaD
```{r, fig.width=8, fig.height=5}
# organ
meta_filter <- meta_filter %>% mutate(organism = factor(organism, levels = meta_filter$organism %>% unique()))
a <- meta_filter %>%
  mutate(Organ = case_when(Organ == 'Eye' ~ 'Eye',
                           TRUE ~ 'Body'),
         Tissue = gsub('\\.|_', ' ', Tissue),
         Tissue = gsub("RPE$","RPE-Choroid", Tissue),
         Tissue = gsub(" and Aorta","", Tissue)) %>% 
  filter(Organ == 'Body') %>% 
  filter(Tissue != 'Brain Choroid Plexus') %>% 
  group_by(Organ, Tissue,organism, study_accession) %>%
  summarise(Count = n()) %>%
  ungroup() %>% 
  mutate(organism = factor(organism, levels = meta_filter$organism %>% unique())) %>% 
  ggplot(aes(x=Tissue,y=Count, fill = organism, group = organism)) +
  ylab('Number of Cells') +
  geom_bar(stat='identity', position = 'dodge') +
  cowplot::theme_cowplot() +
  coord_flip() +
  # scale_y_continuous(breaks = c(2500, 5000, 7500, 10000, 12500, 84000)) +
  # ggbreak::scale_y_break(c(13000, 82000)) +
  scale_fill_discrete(drop=TRUE,
                      limits = levels(meta_filter$organism)) +
  theme(legend.position = "none") +
  scale_y_continuous(expand = c(0, 0)) + 
  ylab('') + xlab('')

aa <- meta_filter %>% 
  filter(Tissue == 'Brain Choroid Plexus') %>% 
  group_by(Organ, Tissue,organism, study_accession) %>%
  summarise(Count = n()) %>%
  ungroup() %>% 
  mutate(organism = factor(organism, levels = meta_filter$organism %>% unique())) %>% 
  ggplot(aes(x=Tissue,y=Count, fill = organism, group = organism)) +
  ylab('Number of Cells') +
  geom_bar(stat='identity', position = 'dodge') +
  cowplot::theme_cowplot() +
  coord_flip() +
  scale_fill_discrete(drop=TRUE,
                      limits = levels(meta_filter$organism)) +
  theme(legend.position = "none") +
  scale_y_continuous(expand = c(0, 0)) + 
  ylab('') + xlab('')

# eye studies
b <-  meta_filter %>%
  mutate(Organ = case_when(Organ == 'Eye' ~ 'Eye',
                           TRUE ~ 'Body'),
         Tissue = gsub('\\.|_', ' ', Tissue),
         Tissue = gsub("RPE$","RPE-Choroid", Tissue),
         Tissue = case_when(Tissue == "Choroid" ~ 'RPE-Choroid',
                            TRUE ~ Tissue),
         Tissue = gsub(" and Aorta","", Tissue)) %>% 
  filter(Organ == 'Eye') %>% 
  group_by(Organ, Tissue,organism) %>%
  summarise(Count = n()) %>%
  ungroup() %>% 
  ggplot(aes(x=Tissue,y=Count, fill = organism, group = organism)) +
  ylab('Number of Cells') +
  geom_bar(stat='identity', position = 'dodge') +
  cowplot::theme_cowplot() +
  coord_flip() +
  theme(legend.position = "none") +
  scale_y_continuous(expand = c(0, 0)) +
  xlab('')

# eye studies samples
c <-  meta_filter %>%
  mutate(Organ = case_when(Organ == 'Eye' ~ 'Eye',
                           TRUE ~ 'Body'),
         Tissue = gsub('\\.|_', ' ', Tissue),
         Tissue = gsub("RPE$","RPE-Choroid", Tissue),
         Tissue = gsub(" and Aorta","", Tissue)) %>% 
  filter(Organ == 'Eye') %>% 
  select(Organ, Tissue,organism, study_accession) %>%
  unique() %>%
  group_by(Organ, Tissue, organism) %>% 
  summarise(Count = n()) %>%
  ungroup() %>% 
  ggplot(aes(x=Tissue,y=Count, fill = organism, group = organism)) +
  ylab('Number of Studies') +
  geom_bar(stat='identity', position = 'dodge') +
  cowplot::theme_cowplot() +
  coord_flip() +
  theme(legend.position = "none") +
  scale_y_continuous(expand = c(0, 0))  + xlab('')

# retina cell types
d <- meta_filter %>% filter(Organ == 'Eye') %>% 
  group_by(CellType_predict, study_accession, organism) %>%
  summarise(Count = n()) %>% filter(Count > 10) %>% ungroup() %>% 
  ggplot(aes(x=CellType_predict,y=Count, fill = organism, group = organism)) +
  ylab('Number of Cells') +
  geom_bar(stat='identity', position = 'dodge') +
  cowplot::theme_cowplot() +
  coord_flip() + 
  scale_y_continuous(expand = c(0, 0)) + xlab('')

a_aa_b <- plot_grid(a,aa,b, ncol = 1, align = 'hv', rel_heights = c(10, 1.5,3.5), labels = 'a')
c_d <- plot_grid(c,d, rel_heights = c(2,10), ncol = 1, align= 'hv', axis = 'l',labels = c('b','c'))
plot_grid(a_aa_b, c_d, ncol = 2, rel_widths = c(1,1.2))
```

@@

## Pan Atlases contain limited retina cell types or lack integration

Tabula Sapiens, MOCA (?), BBrowser (\$\$\$), https://cells.ucsc.edu

## Data difficult to access and cross comparisons impossible

But important as seeing trends trends across different groups strongly demonstrates that 
the signal is biological, not technical.

@@

# Comparison of PLAE to "pan" web apps with retina data
```{r}
feature_table <- cbind(c('Spectacle', 'No' ,'No', ),
                       c('UCSC Cell Browser', 'Yes', 'No', 'https://cells.ucsc.edu' ),
                       c('Tabula Sapiens', 'No', 'N/A', 'https://tabula-sapiens-portal.ds.czbiohub.org'),
                       c('PLAE','Yes','Yes', 'plae.nei.nih.gov'),
                       c('PanglaoDB', 'Yes', 'No'),
                       c('Protein Atlas', 'Yes','No', 'https://www.proteinatlas.org/humanproteome/celltype'))
colnames(feature_table) <- c('Resource Name', 'Independent Datasets', 'Harmonization', 'URL')
```

## PLAE contains over 10 fold more cells and studies than the closest comparable web app

The Spectacle web app also provides a web app to access 87,305 single cell transcriptomes across four publications.
